### 2022-10-30 dissertation/chapters/05-probability.tex
🚧 (chapter5): Conjugates review done

Chapter \ref{chap:hhs} provided the reader with the concept of a \acf{HH}. This dissertation aims to develop a selection \acs{HH} that makes use of probability theory to select the best \ac{HH} to solve a given problem. This chapter aims to provide the reader with the necessary background information on probability theory and statistics. These are big fields and focus is put on the elements that are required to formulate the proposed \acf{BHH}. The remainder of the chapter is structured as follows.

@@

In everyday conversation, the term \textit{probability} is a measure of one's belief in the occurrence of a future event~\cite{ref:wackerly:2014}. Probability is a necessary tool used in many fields including physics, biology, chemistry and computer science. These fields contain many cases that generate observations that can not be predicted with absolute certainty~\cite{ref:wackerly:2014}. Probability can be be inferred and confirmed through past events. These events are referred to as \textit{random} or \textit{stochastic} events. The probability that a certain event $A$ might occur is denoted $P(A)$. Although these random events cannot be predicted with absolute certainty, the relative frequency with which they occur over many trials if often remarkably stable.

Consider flipping an unbiased, fair coin. The coin has two possible outcomes. From this, one can conclude that each side has a $\frac{1}{2}$ or $50\%$ chance of occurring. In statistics, the decimal probability notation is used, where $0 <= P(A) <= 1$. Suppose the fair coin is thrown 10 times, one is not guaranteed to observe $0.5$ heads and $0.5$ tails. There is some probability (although small) that the coin might fall on heads $0/10$ times. The probability of such an event occurring is $0.0009765625$. The \ac{CLT} shows that the normalised sum of events tends toward a normal distribution with a mean value of $0.5$ if the number of events observed $N$, is large~\cite{ref:wackerly:2014}. The larger the value of $N$, the higher the confidence of mean probability and relative frequency of the event. The stable long-term relative frequency by which a random event occurs, provides an intuitive and meaningful measure of belief that a certain event will occur again at some point in the future~\cite{ref:wackerly:2014}.

Probability can also be expressed over multiple random events. For example, again consider flipping a fair, unbiased coin, but this time with a fair, unbiased 6-sided dice. One could consider the probability of observing a certain outcome for the coin together with a certain outcome for the dice. Multiple random events can be considered together, dependently or conditionally. The following sections provides insight into conditional and joints probabilities.

@@

The occurrence of a given random event $A$ can often be dependent on the occurrence of another event $B$. In the field of medicine, an example of this is to calculate the probability of a certain diagnosis of a sick patient given his/her symptoms. The aforementioned dependence is referred to as the conditional probability between two events. The conditional probability is expressed as $P(A \vert B)$ and is read \textit{the probability of $A$ given $B$}. On the contrary, the \textit{unconditional} probability is the probability of an event, not dependent on any other. The conditional probability of $A$ given $B$ can be expressed as is given in Definition \ref{eq:probability:cond_probability} below~\cite{ref:wackerly:2014}.

@@

\begin{definition}[\textbf{Independence of Events}]
      \label{def:probability:cond_probability:independence}
      Two events, $A$ and $B$ are said to be independent of each other if, and only if the following criteria holds:

@@

      otherwise events $A$ and $B$ are said to be \textit{dependent} random events.
\end{definition}

\noindent
Events can also be considered together. The laws of probability for multiple events are presented next.

@@

      \begin{equation*}
            \begin{split}
                  P(A \cap B)
                  &= P(A)P(B \vert A) \\
                  &= P(B)P(A \vert B)
            \end{split}
      \end{equation*}

@@

      \begin{equation*}
            P(A \cap B) = P(A)P(B)
      \end{equation*}
\end{theorem}

@@

      \begin{equation*}
            P(A \cup B) = P(A) + P(B) - P(B \cap A)
      \end{equation*}
\end{theorem}

@@

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.7\textwidth]{images/additive_law_of_probability_proof.png}
      \caption{A Venn-Diagram showing the proof of the additive law of probability for multiple events.}
      \label{fig:probability:multiple_events:additive}
\end{figure}

@@

      \noindent
      Then the collection of sets [$B_{1}$, $B_{2}$, $\dots$, $B_{K}$] is said to be a partition of S.
\end{definition}

\begin{theorem}[\textbf{Bayes' Theorem}]
      \label{th:probability:bayes_theorem:theorem}
      Assume that $[B_{1}, B_{2}, \dots, B_{K}]$ is a partition of $S$ such that $P(B_{k}) > 0$, for $k = 1,2, \dots, K$ then\\
      \\
      \begin{equation*}
            P(B_{k} \vert A) = \frac{P(A \vert B_{k})P(B_{k})}{\sum_{k=1}^{K} P(A \vert B_{k})P(B_{k})}
      \end{equation*}
\end{theorem}

\begin{proof}
      The proof follows directly from the definition of conditional probability as was presented in \ref{sec:probability:cond_probability}. This is shown below

      \begin{equation*}
            \begin{split}
                  P(B_{k} \vert A)
                  &= \frac{P(A \cap B_{k})}{P(A)}\\
                  &= \frac{P(A \vert B_{k})P(B_{k})}{\sum_{k=1}^{K} P(A \vert B_{k})P(B_{k})}
            \end{split}
      \end{equation*}
\end{proof}

@@

\subsection{Beta Distribution}\label{sec:probability:probability_distributions:beta}

The Beta distribution is a family of univariate, continuous, probability distributions over some $x$, with support on the interval $[0,1]$~\cite{ref:wackerly:2014}. It is parameterised by two shape parameters $\alpha > 0$, $\alpha \in \mathbb{R}$ and $\beta > 0$, $\beta \in \mathbb{R}$. The Beta distribution is denoted as $Beta(\alpha, \beta)$. The \ac{PDF} of the Beta distribution is given in Equation~\eqref{eq:probability:probability_distributions:beta:pdf} below.

@@

\noindent
The control parameters $\alpha$ and $\beta$, determine the shape of the distribution. There exists a special case where $\alpha = \beta$. This is referred to as the \textit{symmetric} Beta distribution. In the case where $\alpha = \beta = 1$ the distribution is equivalent to the uniform distribution over all points in its support. The Beta distribution for various values of $\alpha$ and $\beta$, including the symmetric version is presented in Figure \ref{fig:probability:probability_distributions:beta}.

\begin{figure}[htbp]
      \begin{subfigure}{0.49\textwidth}
            \centering
            \includegraphics[width=\textwidth]{images/beta_various.pdf}
            \caption{Beta distribution}
            \label{fig:probability:probability_distributions:beta_normal}
      \end{subfigure}
      \begin{subfigure}{0.49\textwidth}
            \centering
            \includegraphics[width=\textwidth]{images/beta_cumulative.pdf}
            \caption{Cumulative Beta distribution}
            \label{fig:probability:probability_distributions:beta_cumulative}
      \end{subfigure}
      \par\bigskip
      \caption{An illustration of the Beta distribution (left)~\cite{ref:beta:2014} as well as the cumulative Beta distribution (right)~\cite{ref:cumulativebeta:2014} for various values of $\alpha$ and $\beta$ .}
      \label{fig:probability:probability_distributions:beta}
\end{figure}

@@

\subsection{Dirichlet Distribution}
\label{sec:probability:probability_distributions:dirichlet}

The Dirichlet distribution is a family of multivariate continuous probability distributions over some $x$ in $K$ dimensions ~\cite{ref:wackerly:2014}. The Dirichlet distribution is the multivariate generalization of the Beta distribution and is thus sometimes referred to by its alternative name, the Multivariate Beta Distribution. The Dirichlet distribution is parameterised by some vector $\alpha = (\alpha_{1}$,  $\dots, \alpha_{K})$ $\forall_{k=1}^{K} \alpha_{k} > 0$, $\alpha_{k} \in \mathbb{R}$. $\alpha$ is referred to as the concentration parameter. The Dirichlet distribution of order $K \geq 2$ with parameters $\alpha$, denoted $Dir(\alpha)$, has a \ac{PDF} as given in Equation~\eqref{eq:probability:probability_distributions:dirichlet:pdf} below.

@@

\noindent
Importantly, the set $\{x_{k}\}_{k=1}^{K}$ belongs to the standard $K-1$ probability simplex $S$, meaning that $x_{K} = 1 - \sum_{k=1}^{K-1}x_{k}$ with support $\forall_{k=1}^{K} x \in [0,1]$. Under the simplex $S$, this means that the sum over all values of the vector $x$ must be 1. The simplex can thus be rewritten as $\sum_{k=1}^{K}x_{k} = 1$.

Similar to the Beta distribution, $\alpha$ determines the shape of the distribution in $K$ dimensions and thus, there also exist a special case, referred to as the \textit{symmetric} distribution when $\forall_{k=1}^{K} \alpha_{k} = c$, where $c$ is some constant. In the case where $c = 1$, the distribution is referred to as a \textit{flat} distribution and yields the uniform distribution over all points in $S$. The Dirichlet distribution of order $K = 3$, for various values of $\alpha$, including the symmetric version is presented in Figure \ref{fig:probability:probability_distributions:dirichlet}.

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.8\textwidth]{images/dirichlet.pdf}
      \caption{The \acfp{PDF} for the Dirichlet distribution over the 2-simplex. The concentration parameter $\alpha$ is varied. The values of the \acs{PDF} are shown by the color maps with contour lines at equal values as indicated in the color bars~\cite{ref:dirichlet:2020}.}
      \label{fig:probability:probability_distributions:dirichlet}
\end{figure}

@@

\subsection{Bernoulli Distribution}
\label{sec:probability:probability_distributions:bernoulli}

The Bernoulli distribution is a discrete probability distribution over some random variable $x$ that takes the value of $1$ with probability $\theta$ and $0$ with probability $1-\theta$ ~\cite{ref:wackerly:2014}. It is denoted as $Ber(\theta)$ with support $x \in \{0, 1\}$. In probability theory, the Bernoulli distribution is often used to explain the possible outcomes of a single experiment that asks a \textit{yes-no} question such as flipping a coin. The outcome of such an experiment is a boolean value. The Bernoulli distribution has a \ac{PMF} as given below in Equation~\eqref{eq:probability:probability_distributions:bernoulli:pmf}.

@@

The above equation can also be expressed as show in Equation~\eqref{eq:probability:probability_distributions:bernoulli:alt} below.

@@

The mean of the Bernoulli distribution approaches $\theta$ over many samples according to the \ac{CLT}~\cite{ref:grinstead:1997}. Figure \ref{fig:probability:probability_distributions:bernoulli:coin} shows a simulation of a fair-coin flipping simulation. One can see how the mean converges to 0.5 for various sample sizes.

\begin{figure}[htbp]
      \begin{subfigure}{0.49\textwidth}
            \includegraphics[width=\textwidth]{images/coin_flip_samples_100.png}
            \caption{100 Samples}
            \label{fig:probability:probability_distributions:bernoulli:coin_100}
      \end{subfigure}
      \begin{subfigure}{0.49\textwidth}
            \includegraphics[width=\textwidth]{images/coin_flip_samples_10000.png}
            \caption{1000 Samples}
            \label{fig:probability:probability_distributions:bernoulli:coin_1000}
      \end{subfigure}
      \par\bigskip
      \caption{An illustration of the the coin flip situation for different sample sizes that show the convergence of the mean as per the \ac{CLT}.}
      \label{fig:probability:probability_distributions:bernoulli:coin}
\end{figure}

The expected value of the distribution is thus given according to Equation~\eqref{eq:probability:probability_distributions:bernoulli:expected_value} below.

@@

while the mode of the distribution is given in Equation~\eqref{eq:probability:probability_distributions:bernoulli:mode} below.

@@

\subsection{Binomial Distribution}
\label{sec:probability:probability_distributions:bin}

The Binomial distribution is a discrete probability distribution over a random variable $x$ taking on a number of successes in $N$ sequential independent experiments that each ask a \textit{yes-no} question~\cite{ref:wackerly:2014}. The probability of a single independent experiment yielding a success is given as $\theta$ and the Binomial distribution is denoted as $Bin(N, \theta)$ with support $x \in \{0, 1, \dots, N\}$.  It should be noted that the Binomial distribution is the extension of the Bernoulli distribution over $N$ independent sequential experiments and thus, each experiment also yields some boolean outcome. When $N=1$, the experiment is referred to as a Bernoulli trial and the distribution is just a Bernoulli distribution. When $N > 1$, the sequence of outcomes is referred to as a Bernoulli process. The \ac{PMF} of the Binomial distribution is given in Equation~\eqref{eq:probability:probability_distributions:binomial:pmf} below.

@@

Similar to the Bernoulli distribution, the mean of the Binomial distribution is $N\theta$ given the \ac{CLT} as was shown in Figure \ref{fig:probability:probability_distributions:bernoulli:coin}. The expected value of the Binomial distribution is thus given in Equation~\eqref{eq:probability:probability_distributions:binomial:expected_value} below.

@@

The mode of the distribution is given in Equation~\eqref{eq:probability:probability_distributions:binomial:mode} below.

@@

\subsection{Categorical Distribution}
\label{sec:probability:probability_distributions:categorical}

The Categorical distribution is a discrete probability distribution over some random variable $x$, taking on any one of $K$ possible categories~\cite{ref:wackerly:2014}. There is no innate underlying ordering to these categories and therefore, for simplicity, each category is assigned a numerical representative value such that $k = (1, \dots, K)$. The probabilities for all outcomes is given by the probability vector $\theta = (\theta_{1}, \dots, \theta_{K})$.  This means that the probability $P(x=k)=\theta_{k}$ has support $x \in \{1, \dots, K\}$. The Categorical distribution, denoted $Cat(\theta)$ is the generalization of the Bernoulli distribution and is sometimes referred to it by its alternative names, the generalised Bernoulli Distribution or the Multinoulli distribution. In probability theory, the Categorical distribution is often used to explain the outcome of a single experiment with more than two possible outcomes such as rolling a die~\cite{ref:wackerly:2014}. The \ac{PMF} of the Categorical distribution is given in Equation~\eqref{eq:probability:probability_distributions:categorical:pmf} below.

@@

where $[x = k]$ is the Iversion Bracket~\cite{ref:iverson:1962}, yielding 1 if $x = k$ and 0 otherwise. From this, one can conclude that for a given class $k$, the categorical distribution simply yields $\theta_{k}$ as is shown in Equation~\eqref{eq:probability:probability_distributions:categorical:pmf_k} below.

@@

The random variable $x$ can also be encoded in binary format, yielding a vector $x = (x_{1}, \dots, x_{K})$ of Bernoulli distributions such that the support is $\forall_{k=1}^{K} x_{k} \in \{0, 1\}$. Importantly, if the outcome of the random event is of category $k$, then $x_{k} = 1$ and $\forall_{j=1}^{K} x_{j} = 0, j \neq k$ so that the standard $K-1$ probability simplex $S$ still holds. The \ac{PMF} of the Categorical distribution can then be rewritten as follows in Equation~\eqref{eq:probability:probability_distributions:categorical:pmf_alt} below.

@@

where $\mathbbm{1}(x_{k})$ is the \textit{indicator function}, yielding 1 if $x_{k} = 1$ and 0 otherwise.

Since there is no innate order to the underlying categories, the mean of the distribution does not yield any relevant information. The mode of the distribution is given in Equation~\eqref{eq:probability:probability_distributions:categorical:mode} below.

@@

The Multinomial distribution is a discrete probability distribution over some random variable $x = (x_{1}, \dots\, x_{K})$ that takes on the counts for each occurrence of $K$ possible classes in $N$ independent trials~\cite{ref:wackerly:2014}. The probabilities for all possible outcomes in a single trial is given by the probability vector $\theta = (\theta_{1}, \dots, \theta_{K})$. The Multinomial distribution, denoted $Mul(N, K, \theta)$, is thus the generalization of the Binomial distribution to $K$ dimensions. Consider the following special cases.

\begin{itemize}
      \item When $K$ is 2 and $N = 1$, the Multinomial distribution is the Bernoulli distribution.
      \item When $K$ is 2 and $N > 1$, the Multinomial distribution is the Binomial distribution.
      \item When $K > 2$ and $N = 1$, the Multinomial distribution is the Categorical distribution.
\end{itemize}

The support for the Multinomial is $\forall_{i=1}^{K} x_{k} \in \{1, \dots, N\}, \sum_{k=1}^{K}x_{k} = N$ and the \ac{PMF} for the Multinomial distribution is given below in Equation~\eqref{eq:probability:probability_distributions:multinomial:pmf}.

@@

Similar to the Categorical distribution, the random variable $x$ can also be encoded in binary format, yielding an $N \times K$ matrix $X$ of Bernoulli distributions. The support is then given as $X \in \{0, 1\}^{N \times K}, \forall_{i=1}^{N}\sum_{k=1}^{K} x_{i,k} = 1$ so that the standard $K-1$ probability simplex $S$ still holds for each trial. The \ac{PMF} of the Multinomial distribution can then be rewritten as is presented in Equation~\eqref{eq:probability:probability_distributions:multinomial:pmf_alt}

@@

where $N_{k}$ is a summary variable denoting the number of times a category $k$ occurs over all trials in $N$.

>>>>

Chapter \ref{chap:hhs} provided the reader with the concept of a \acf{HH}. This dissertation aims to develop a selection \acs{HH} that makes use of probability theory to select the best \acs{HH} to solve a given problem. This chapter aims to provide the reader with the necessary background information on probability theory and statistics. These are big fields and focus is put on the elements that are required to formulate the proposed \acf{BHH}. The remainder of the chapter is structured as follows.

@@

In everyday conversation, the term \textit{probability} is a measure of one's belief in the occurrence of a future event~\cite{ref:wackerly:2014}. Probability is a necessary tool used in many fields including physics, biology, chemistry and computer science. These fields contain many cases that generate observations that cannot be predicted with absolute certainty~\cite{ref:wackerly:2014}. Probability can be be inferred and confirmed through past events. These events are referred to as \textit{random} or \textit{stochastic} events. The probability that a certain event $A$ might occur is denoted $P(A)$. Although these random events cannot be predicted with absolute certainty, the relative frequency with which they occur over many trials, is often remarkably stable.

Consider flipping an unbiased, fair coin. The coin has two possible outcomes. From this, one can conclude that each side has a $\frac{1}{2}$ or $50\%$ chance of occurring. In statistics, the decimal probability notation is used, where $0 <= P(A) <= 1$. Suppose the fair coin is thrown 10 times, one is not guaranteed to observe $0.5$ heads and $0.5$ tails. There is some probability (although small) that the coin might fall on heads $0/10$ times. The probability of such an event occurring is $0.0009765625$. In the coin flip example, the \acf{CLT} shows that the normalised sum of events tends toward a normal distribution with a mean value of $0.5$ if the number of events observed $N$, is large~\cite{ref:wackerly:2014}. The larger the value of $N$, the higher the confidence of mean probability and relative frequency of the event. The stable long-term relative frequency by which a random event occurs, provides an intuitive and meaningful measure of belief that a certain event will occur again at some point in the future~\cite{ref:wackerly:2014}.

Probability can also be expressed over multiple random events. Multiple random events can be considered together, dependently or conditionally. The following sections provides insight into conditional and joints probabilities of multiple random events.

@@

The occurrence of a given random event $A$ can often be conditional on the occurrence of another event $B$. In the field of medicine, an example of this is to calculate the probability of a certain diagnosis of a sick patient given his/her symptoms. The aforementioned relationship between events is referred to as the conditional probability between two events. The conditional probability is expressed as $P(A \vert B)$ and is read \textit{the probability of $A$ given $B$}. On the contrary, the \textit{unconditional} probability is the probability of an event, not dependent on any other. The conditional probability of $A$ given $B$ can be expressed as is given in Definition \ref{eq:probability:cond_probability} below~\cite{ref:wackerly:2014}.

@@

\begin{definition}[\textbf{Independence of Events}]
      \label{def:probability:cond_probability:independence}
      Two events, $A$ and $B$ are said to be independent of each other if, and only if the following criteria hold:

@@

      \noindent
      otherwise events $A$ and $B$ are said to be \textit{dependent} random events.
\end{definition}

@@

      \begin{equation}
            \begin{split}
                  P(A \cap B)
                  &= P(A)P(B \vert A) \\
                  &= P(B)P(A \vert B)
            \end{split}
      \end{equation}

@@

      \begin{equation}
            P(A \cap B) = P(A)P(B)
      \end{equation}
\end{theorem}

@@

      \begin{equation}
            P(A \cup B) = P(A) + P(B) - P(B \cap A)
      \end{equation}
\end{theorem}

@@

\begin{figure}[htbp]
      \centering
      \includegraphics[width=0.7\textwidth]{images/additive_law_of_probability_proof.pdf}
      \caption{A Venn-Diagram showing the proof of the additive law of probability for multiple events.}
      \label{fig:probability:multiple_events:additive}
\end{figure}

@@

      \noindent
      Then the collection of sets $\{B_{1}$, $B_{2}$, $\dots$, $B_{K}\}$ is said to be a partition of S, the union of mutually exclusive subsets.
\end{definition}

\begin{theorem}[\textbf{Bayes' Theorem}]
      \label{th:probability:bayes_theorem:theorem}
      Assume that $[B_{1}, B_{2}, \dots, B_{K}]$ is a partition of $S$ such that $P(B_{i}) > 0$, for $i = 1,2, \dots, K$ then\\
      \\
      \begin{equation}
            P(B_{j} \vert A) = \frac{P(A \vert B_{j})P(B_{j})}{\sum_{i=1}^{K} P(A \vert B_{i})P(B_{i})}
      \end{equation}
\end{theorem}

\begin{proof}
      The proof follows from the definition of conditional probability as was presented in \ref{sec:probability:cond_probability}.

      \begin{equation}
            \begin{split}
                  P(B_{j} \vert A)
                  &= \frac{P(A \cap B_{j})}{P(A)}\\
                  &= \frac{P(A \vert B_{j})P(B_{j})}{\sum_{i=1}^{K} P(A \vert B_{i})P(B_{i})}
            \end{split}
      \end{equation}
\end{proof}

@@

\subsection{Beta probability distribution}\label{sec:probability:probability_distributions:beta}

The \index{Beta probability distribution}Beta probability distribution is a family of univariate, continuous, probability distributions over some $x$, with support on the interval $[0,1]$~\cite{ref:wackerly:2014}. It is parameterised by two shape parameters $\alpha > 0$, $\alpha \in \mathbb{R}$ and $\beta > 0$, $\beta \in \mathbb{R}$. The \index{Beta probability distribution}Beta probability distribution is denoted as $Beta(\alpha, \beta)$. The \acf{PDF} of the \index{Beta probability distribution}Beta probability distribution is given in Equation~\eqref{eq:probability:probability_distributions:beta:pdf} below.

@@

\noindent
The control parameters $\alpha$ and $\beta$, determine the shape of the distribution. There exists a special case where $\alpha = \beta$. This is referred to as the \textit{symmetric} \index{Beta probability distribution}Beta probability distribution. In the case where $\alpha = \beta = 1$ the distribution is equivalent to the uniform distribution over all points in its support. The \index{Beta probability distribution}Beta probability distribution for various values of $\alpha$ and $\beta$, including the symmetric version is presented in Figure \ref{fig:probability:probability_distributions:beta}.

\begin{figure}[htbp]
      \begin{subfigure}{0.49\textwidth}
            \centering
            \includegraphics[width=\textwidth]{images/beta_various.pdf}
            \caption{\index{Beta probability distribution}Beta probability distribution}
            \label{fig:probability:probability_distributions:beta_normal}
      \end{subfigure}
      \begin{subfigure}{0.49\textwidth}
            \centering
            \includegraphics[width=\textwidth]{images/beta_cumulative.pdf}
            \caption{Cumulative \index{Beta probability distribution}Beta probability distribution}
            \label{fig:probability:probability_distributions:beta_cumulative}
      \end{subfigure}
      \par\bigskip
      \caption{An illustration of the \index{Beta probability distribution}Beta probability distribution (left)~\cite{ref:beta:2014} as well as the cumulative \index{Beta probability distribution}Beta probability distribution (right)~\cite{ref:cumulativebeta:2014} for various values of $\alpha$ and $\beta$ .}
      \label{fig:probability:probability_distributions:beta}
\end{figure}

@@

\subsection{Dirichlet probability distribution}\label{sec:probability:probability_distributions:dirichlet}

The \index{Dirichlet probability distribution}Dirichlet probability distribution is a family of multivariate continuous probability distributions over some $x$ in $K$ dimensions~\cite{ref:wackerly:2014}. The \index{Dirichlet probability distribution}Dirichlet probability distribution is the multivariate generalization of the \index{Beta probability distribution}Beta probability distribution and is thus sometimes referred to by its alternative name, the multivariate Beta probability distribution. The \index{Dirichlet probability distribution}Dirichlet probability distribution is parameterised by some vector $\vec{\alpha} = (\alpha_{1}$,  $\dots, \alpha_{K})$ $\forall_{k=1}^{K} \alpha_{k} > 0$, $\alpha_{k} \in \mathbb{R}$. $\alpha$ is referred to as the concentration parameter. The \index{Dirichlet probability distribution}Dirichlet probability distribution of order $K \geq 2$ with parameters $\alpha$, denoted $Dir(\alpha)$, has a \acs{PDF} as is given in Equation~\eqref{eq:probability:probability_distributions:dirichlet:pdf} below.

@@

\noindent
Importantly, the set $\{x_{k}\}_{k=1}^{K}$ belongs to the standard $K-1$ probability simplex $S$, meaning that $x_{K} = 1 - \sum_{k=1}^{K-1}x_{k}$ with support $\forall_{k=1}^{K} x \in [0,1]$. Under the simplex $S$, this means that the sum over all values of the vector $\vec{x}$ must be 1. The simplex can thus be rewritten as $\sum_{k=1}^{K}x_{k} = 1$.

Similar to the \index{Beta probability distribution}Beta probability distribution, $\alpha$ determines the shape of the distribution in $K$ dimensions and thus, there also exist a special case, referred to as the \textit{symmetric} distribution when $\forall_{k=1}^{K} \alpha_{k} = c$, where $c$ is some constant. In the case where $c = 1$, the distribution is referred to as a \textit{flat} distribution and yields the uniform distribution over all points in $S$. The \index{Dirichlet probability distribution}Dirichlet probability distribution of order $K = 3$, for various values of $\alpha$, including the symmetric version is presented in Figure \ref{fig:probability:probability_distributions:dirichlet}.

\begin{figure}[htb]
      \centering
      \includegraphics[width=0.9\textwidth]{images/dirichlet.pdf}
      \caption{The \acfp{PDF} for the \index{Dirichlet probability distribution}Dirichlet probability distribution over the 2-simplex. The concentration parameter $\alpha$ is varied. The values of the \acs{PDF} are shown by the color maps with contour lines at equal values as indicated in the color bars~\cite{ref:dirichlet:2020}.}
      \label{fig:probability:probability_distributions:dirichlet}
\end{figure}

@@

\subsection{Bernoulli probability distribution}\label{sec:probability:probability_distributions:bernoulli}

The \index{Bernoulli probability distribution}Bernoulli probability distribution is a discrete probability distribution over some random variable $x$ that takes the value of $1$ with probability $\theta$ and $0$ with probability $1-\theta$ ~\cite{ref:wackerly:2014}. The \index{Bernoulli probability distribution}Bernoulli probability distribution is denoted as $Ber(\theta)$ with support $x \in \{0, 1\}$. In probability theory, the \index{Bernoulli probability distribution}Bernoulli probability distribution is often used to explain the possible outcomes of a single experiment that asks a \textit{yes-no} question such as flipping a coin. The outcome of such an experiment is a boolean value. The \index{Bernoulli probability distribution}Bernoulli probability distribution has a \acf{PMF} as given below in Equation~\eqref{eq:probability:probability_distributions:bernoulli:pmf}.

@@

\noindent
Equation~\eqref{eq:probability:probability_distributions:bernoulli:pmf} above, can also be expressed as show in Equation~\eqref{eq:probability:probability_distributions:bernoulli:alt} below.

@@

\noindent
The mean of the \index{Bernoulli probability distribution}Bernoulli probability distribution approaches $\theta$ over many samples according to the \acs{CLT}~\cite{ref:grinstead:1997}. Figure \ref{fig:probability:probability_distributions:bernoulli:coin} shows a simulation of a fair-coin flipping simulation. The mean converges to 0.5 for various sample sizes.

\begin{figure}[htbp]
      \centering
      \begin{subfigure}{0.49\textwidth}
            \centering
            \includegraphics[width=\textwidth]{images/coin_flip_samples_100.pdf}
            \caption{100 Samples}
            \label{fig:probability:probability_distributions:bernoulli:coin_100}
      \end{subfigure}
      \begin{subfigure}{0.49\textwidth}
            \centering
            \includegraphics[width=\textwidth]{images/coin_flip_samples_10000.pdf}
            \caption{10000 Samples}
            \label{fig:probability:probability_distributions:bernoulli:coin_10000}
      \end{subfigure}
      \par\bigskip
      \caption{An illustration of the the coin flip situation for different sample sizes that show the convergence of the mean as per the \acf{CLT}.}
      \label{fig:probability:probability_distributions:bernoulli:coin}
\end{figure}

\noindent
The expected value of the distribution is thus given according to Equation~\eqref{eq:probability:probability_distributions:bernoulli:expected_value} below.

@@

\noindent
while the mode of the distribution is given in Equation~\eqref{eq:probability:probability_distributions:bernoulli:mode} below.

@@

\subsection{Binomial Distribution}\label{sec:probability:probability_distributions:bin}

The \index{Binomial distribution}Binomial distribution is a discrete probability distribution over a random variable $x$ taking on a number of successes in $N$ sequential independent experiments that each ask a \textit{yes-no} question~\cite{ref:wackerly:2014}. The probability of a single independent experiment yielding a success, is given as $\theta$ and the \index{Binomial distribution}Binomial distribution is denoted as $Bin(N, \theta)$ with support $x \in \{0, 1, \dots, N\}$.  It should be noted that the \index{Binomial distribution}Binomial distribution is the extension of the \index{Bernoulli probability distribution}Bernoulli probability distribution over $N$ independent sequential experiments and thus, each experiment also yields some boolean outcome. When $N=1$, the experiment is referred to as a Bernoulli trial and the distribution is just a \index{Bernoulli probability distribution}Bernoulli probability distribution. When $N > 1$, the sequence of outcomes is referred to as a Bernoulli process. The \acf{PMF} of the \index{Binomial distribution}Binomial distribution is given in Equation~\eqref{eq:probability:probability_distributions:binomial:pmf} below.

@@

\noindent
Similar to the \index{Bernoulli probability distribution}Bernoulli probability distribution, the mean of the \index{Binomial distribution}Binomial distribution is $N\theta$ given the \acs{CLT} as was shown in Figure \ref{fig:probability:probability_distributions:bernoulli:coin}. The expected value of the \index{Binomial distribution}Binomial distribution is thus given in Equation~\eqref{eq:probability:probability_distributions:binomial:expected_value} below.

@@

\noindent
The mode of the distribution is given in Equation~\eqref{eq:probability:probability_distributions:binomial:mode} below.

@@

\subsection{Categorical Probability Distribution}\label{sec:probability:probability_distributions:categorical}

The \index{Categorical probability distribution}Categorical probability distribution is a discrete probability distribution over some random variable $x$, taking on any one of $K$ possible categories~\cite{ref:wackerly:2014}. There is no ordering to these categories and therefore, for simplicity, each category is assigned a numerical representative value such that $k = (1, 2, \dots, K)$. The probabilities for all outcomes is given by the probability vector $\vec{\theta} = (\theta_{1}, \dots, \theta_{K})$.  This means that the probability $P(x=k)=\theta_{k}$, has support $x \in \{1, \dots, K\}$. The \index{Categorical probability distribution}Categorical probability distribution, denoted $Cat(\theta)$, is the generalization of the \index{Bernoulli probability distribution}Bernoulli probability distribution and is sometimes referred to it by its alternative names, the generalised \index{Bernoulli probability distribution}Bernoulli probability distribution or the Multinoulli probability distribution. In probability theory, the \index{Categorical probability distribution}Categorical probability distribution is often used to explain the outcome of a single experiment with more than two possible outcomes such as rolling a die~\cite{ref:wackerly:2014}. The \acs{PMF} of the \index{Categorical probability distribution}Categorical probability distribution is given in Equation~\eqref{eq:probability:probability_distributions:categorical:pmf} below.

@@

\noindent
In Equation~\eqref{eq:probability:probability_distributions:categorical:pmf} above, $[x = k]$ is the \textit{Iversion Bracket}~\cite{ref:iverson:1962}, yielding 1 if $x = k$ and 0 otherwise. Given class $k$, the categorical distribution simply yields $\theta_{k}$ as is shown in Equation~\eqref{eq:probability:probability_distributions:categorical:pmf_k} below.

@@

\noindent
The random variable $x$ can also be encoded in binary format, yielding a vector $\vec{x} = (x_{1}, \dots, x_{K})$ of \index{Bernoulli probability distribution}Bernoulli probability distributions such that the support is $\forall_{k=1}^{K} x_{k} \in \{0, 1\}$. Importantly, if the outcome of the random event is of category $k$, then $x_{k} = 1$ and $\forall_{j=1}^{K} x_{j} = 0, j \neq k$ so that the standard $K-1$ probability simplex $S$ still holds. The \acs{PMF} of the \index{Categorical probability distribution}Categorical probability distribution can then be rewritten as follows in Equation~\eqref{eq:probability:probability_distributions:categorical:pmf_alt} below.

@@

\noindent
In Equation \ref{eq:probability:probability_distributions:categorical:pmf_alt} above, $\mathbbm{1}(x_{k})$ is the \textit{indicator function}, yielding 1 if $x_{k} = 1$ and 0 otherwise.

Since there is no order to the underlying categories, the mean of the distribution does not yield any relevant information. The mode of the distribution is given in Equation~\eqref{eq:probability:probability_distributions:categorical:mode} below.

@@

The \index{Multinomial probability distribution}Multinomial probability distribution is a discrete probability distribution over some random variable $x = (x_{1}, \dots\, x_{K})$ that takes on the counts for each occurrence of $K$ possible classes in $N$ independent trials~\cite{ref:wackerly:2014}. The probabilities for all possible outcomes in a single trial is given by the probability vector $\vec{\theta} = (\theta_{1}, \dots, \theta_{K})$. The \index{Multinomial probability distribution}Multinomial probability distribution, denoted $Mul(N, K, \theta)$, is thus the generalization of the \index{Binomial distribution}Binomial distribution to $K$ dimensions. Consider the following special cases.

\begin{itemize}
      \item When $K$ is 2 and $N = 1$, the \index{Multinomial probability distribution}Multinomial probability distribution is the \index{Bernoulli probability distribution}Bernoulli probability distribution.
      \item When $K$ is 2 and $N > 1$, the \index{Multinomial probability distribution}Multinomial probability distribution is the \index{Binomial distribution}Binomial distribution.
      \item When $K > 2$ and $N = 1$, the \index{Multinomial probability distribution}Multinomial probability distribution is the \index{Categorical probability distribution}Categorical probability distribution.
\end{itemize}

\noindent
The support for the Multinomial is $\forall_{i=1}^{K} x_{k} \in \{1, \dots, N\}, \sum_{k=1}^{K}x_{k} = N$ and the \acs{PMF} for the \index{Multinomial probability distribution}Multinomial probability distribution is given below in Equation~\eqref{eq:probability:probability_distributions:multinomial:pmf}.

@@

\noindent
Similar to the \index{Categorical probability distribution}Categorical probability distribution, the random variable $x$ can also be encoded in binary format, yielding an $N \times K$ matrix $X$ of \index{Bernoulli probability distribution}Bernoulli probability distributions. The support is then given as $X \in \{0, 1\}^{N \times K}, \forall_{i=1}^{N}\sum_{k=1}^{K} x_{i,k} = 1$ so that the standard $K-1$ probability simplex $S$ still holds for each trial. The \acs{PMF} of the \index{Multinomial probability distribution}Multinomial probability distribution can then be rewritten as is presented in Equation~\eqref{eq:probability:probability_distributions:multinomial:pmf_alt}

@@

\noindent
In Equation~\eqref{eq:probability:probability_distributions:multinomial:pmf_alt} above, $N_{k}$ is a summary variable denoting the number of times a category $k$ occurs over all trials in $N$.
